cardinalR: Generating interesting high-dimensional data structures

### 2025-05-05 paper-cardinalR.Rmd
Added hemisphere

@@

@@

>>>>

The `gen_clusteredspheres(n, k, p, r, loc)` function generates a synthetic dataset of $n_1 + k \cdot n_2$ observations in $p$-dimensional space, consisting of one large sphere of radius $r_1$ and $k$ smaller spheres of radius $r_2$, each centered at a different random location.

#### Geometric Structure

- A **large uniform sphere** centered at the origin:
  - \(n_1\) points are sampled **uniformly on the surface** of a \(p\)-dimensional sphere of radius \(r_1\).
  - Generated using `gen_unifsphere(n_1, p, r_1)`.

- \(k\) **smaller uniform spheres**, each with:
  - \(n_2\) points sampled uniformly on a sphere of radius \(r_2\),
  - Positioned at different random locations in \(p\)-space:
    - Each center is drawn from \( \mathcal{N}(0, \texttt{loc}^2 I_p) \),
    - Ensures separation between clusters.

All points on all spheres are generated using the standard hyperspherical method:

- For each point:
  - \(u \sim U(-1, 1)\), the cosine of the polar angle,
  - \(\theta \sim U(0, 2\pi)\), the azimuthal angle (for 3D),
  - Generalized to \(p\)-dimensions to uniformly sample on the sphere’s surface.

Each observation is classified by cluster, with labels such as "big" for the large central sphere and "small_1" to "small_k" for the smaller spheres.

@@

The `gen_hemisphere(n, p)` function generates a $p$-dimensional dataset of $n$ observations distributed approximately uniformly on a $4\text{-}D$ hemisphere, optionally extended with Gaussian noise in additional dimensions when $p > 4$.

Each observation is situated on a restricted $4\text{-}D$ spherical surface, defined by spherical coordinates. The azimuthal angle $\theta_1 \sim U(0, \pi)$ in the $(x_1, x_2)$ plane, while the elevation angle $\theta_2 \sim U(0, \pi)$ in the $(x_2, x_3)$ plane. Additionally, $\theta_3 \sim U(0, \pi/2)$ in the $(x_3, x_4)$ plane, ensuring that the points remain restricted to a hemisphere.

The coordinates are transformed into $4\text{-}D$ Cartesian space:

$$
X_1 = \sin(\theta_1) \cdot \cos(\theta_2),
$$
$$
X_2 = \sin(\theta_1) \cdot \sin(\theta_2),
$$
$$
X_3 = \cos(\theta_1) \cdot \cos(\theta_3),
$$
$$
X_4 = \cos(\theta_1) \cdot \sin(\theta_3).
$$

This produces points on one side of a $4\text{-}D$ unit sphere, effectively generating a $4\text{-}D$ hemisphere.

For $p > 4$, additional dimensions $X_5$ to $X_p$ are added as low-variance Gaussian noise:

$$
X_j \sim \mathcal{N}(0, 0.05^2),\text{ for }j = 5, \dots, p.
$$

These dimensions maintain the spherical structure while simulating embedding into higher-dimensional space with small orthogonal perturbations.

@@

- clusteredsphere: This structure allows for cluster separation on curved manifolds in high-dimensional space and can be used to test the ability of NLDR methods and clustering algorithms to detect spherical clusters of different sizes and separations.
